---
title: "Qualitative text analysis with local LLMs: Part III"
date: 2025-01-28
categories: ['llm', 'R']
description: |
    In this third part of a three part note on qualitative text analysis with LLMs, 
    I discuss some practical issues concerning using local LLMs, especially the 
    computational power requirements.
---

```{r}
#| message: false
#| echo: false
library(tidyverse)
```

In [Part I](part1.html) of this note, we described how to install and use local LLMs, and in [Part II](part2.html), we described how to use a local LLM to rate free-text survey responses like a human rater.
Here, we want to very briefly consider some practical issues concerning using local LLMs for text analysis.
In particular I consider the computational power needed for local LLMs.

# Computational power

Thus far, we have not directly discussed the computational power requirements for running local LLMs.
I have not done anything like a comprehensive analysis, I have just speed tested Llama3.3 doing one task on two devices, one that has a CPU only and another that has a good GPU.

* CPU only device: AMD Ryzen 9 7900x 12-Core, 94GB of DDR5 RAM
* GPU device: AMD Ryzen Threadripper PRO 3995WX 64-Cores, 512GB DDR4 RAM, with a RTX A6000 GPU (10752 CUDA cores, 48GB of VRAM).

Of course, these two devices differ not simply by one having a GPU.
The Threadripper also has many more CPU cores and considerably more RAM too.
To be able to identify the speed-up from the GPU per se, we will also test the Threadripper with the GPU disabled.

It should be pointed out the the Ryzen 9, although a bit out of date now, is still a relatively high-end desktop CPU as of late 2024.
The Threadripper is a bit of date now, but a few years ago would have been on the high-end for a server or scientific workstation CPU.
The RTX A6000, also a bit out of date now, is still a relatively high-end GPU.

For the tests, I ran the following `speed_test.R` script, which is based on the task described in [Part II](part2.html):

```{r}
#| eval: false
# speed_test.R
free_text <- "
I didn't really want people to be afraid of me. I had taken a test and 
I knew I wasn't positive for Covid, so I was just worried that people would think
it was false and that they'd avoid me or get mad at me for attending when I was sick
"

instructions <- 'We have collected free response data from students at the University of Michigan
and healthcare workers within Michigan Medicine.
Your job will be to read through these free responses about why participants said they were motivated 
to hide signs of infectious illness and indicate where they fall on a number of different variables.

Does the participant mention motivations for concealment that were more related to the self or more 
related to others? Examples of self motivation include not wanting to miss out on in-person things
like work or class, not wanting others to judge them, not wanting others to avoid them.
Examples of other motivation include not wanting to worry other people, not wanting to burden
others by missing a work shift

Coding scheme: please put a "1" if it is a self motivation, a "2" if it is an other motivation,
and a "0" if it is neither/unclear.
Note: It is possible for a response to mention both self and other reasons, so it is ok for there
to be both a 1 and a 2 for the same response, i.e. "1,2".

Before answering, explain your reasoning step by step, using example phrases or words.
Then provide the final answer.

Your final answer should be either "0" or "1" or "2" or "1,2".

Format your response as a JSON object literal with keys "reasoning" and "answer".
The value of "reasoning" is your step by step reasoning, using phrases or words.
The value of "answer" is the numerical score ("0", "1", "2", or "1,2") that you assign
to the free response text.

And example of a JSON object literal response is this:

  {"reasoning": "The text shows many elements of a self motivation.",
  "answer": "1"}

Please make sure that there is an opening (left) and closing (right) curly brace in what you return.
If not, this is not a proper JSON object literal.
'

client <- ellmer::chat_ollama(model = "llama3.3", system_prompt = instructions)
results <- client$chat(free_text)
```

```{r}
#| echo: false
# six runs of the `speed_test.R on AMD Ryzen 9 7900X 12-Core Processor, 94GB RAM
cpu_device <- c(60 + 35, 2*60 + 9, 60 + 54, 60 + 39, 2 * 60 + 5, 60 + 33)
# six runs of `speed_test.R` on AMD Ryzen Threadripper PRO 3995WX 64-Cores, 512GB (64 x 8 DDR4), RTX A6000 (10752 CUDA cores; 10,752 CUDA cores, 48GB of GDDR6 VRAM, and a 768 GB/s memory bandwidth)
gpu_device <- c(10.4, 7.0, 10.5, 8.0, 8.1, 7.9)
# six runs of `speed_test.R` on the same Threadripper but now ollama-cuda was uninstalled (pacman -R ollama-cuda on Arch Linux) and so only the CPU was used
gpu_device_cpu_only <- c(40.1, 38.3, 44.6, 48, 43.3, 53.2)

# average comparison on Ryzen 9 with Threadripper using the GPU
avg_ratio <- mean(cpu_device) / mean(gpu_device)
# average ratio between any one run on the CPU only device and any one on the GPU only device
N <- 1e4
avg_ratio2 <- mean(sample(cpu_device, size = N, replace = TRUE) / sample(gpu_device, size = N, replace = TRUE))
# average comparison on Threadripper with or without the GPU
avg_ratio_thread <- mean(gpu_device_cpu_only) / mean(gpu_device)
```

This test code was run six times on Ryzen 9 device, six times on the Threadripper with the GPU disabled, and six times on the Threadripper with the GPU enabled.
The wall clock times for each run, in seconds, are as follows:
```{r}
#| echo: false
tibble(`Ryzen 9` = cpu_device, `Threadripper without GPU` = gpu_device_cpu_only, `Threadripper with GPU` = gpu_device) |> 
    mutate(run = seq(n()), .before = everything()) |> 
    knitr::kable(caption = 'Running time (seconds) of `speed_test.R`') 
```

The Ryzen 9 is on average `r round(mean(cpu_device)/mean(gpu_device_cpu_only), 1)` slower than the Threadripper using only the CPU.
The Threadripper using only the CPU only is on average `r round(mean(gpu_device_cpu_only)/mean(gpu_device), 1)` slower than the Threadripper using the GPU.
The Ryzen 9 is on average `r round(mean(cpu_device)/mean(gpu_device), 1)` slower than the Threadripper using the GPU.

Although this is a very minimal investigation, I think it is still informative.
In particular, it seems to tell us that high-end desktop devices alone are probably not sufficient for serious work with local LLMs, and high-end GPUs are probably necessary.

If we treat the Ryzen 9 as a high-end desktop device (as of around 2024), this test shows that a high-end desktop is not sufficient for using local LLMs for serious work unless it is on a small scale.
The task described in [Part II](part2.html) essentially involves running the `speed_test.R` code $4 \times 409 \times 10 = `r format(4 * 409 * 10, scientific=FALSE)`$ times (409 different texts; 4 separate ratings for each; 10 repetitions).
This means that it will take approximately `r format(round(mean(cpu_device) * 4 * 409 * 10 / (60 * 60 * 24)), scientific=FALSE)` days to complete.
That, I presume, would stretch anyone's patience, especially given that in any given analysis, we are likely to want to try multiple different prompts and other changes.
By contrast to the Ryzen 9, the Threadripper using its GPU would take around `r format(round(mean(gpu_device) * 4 * 409 * 10 / (60 * 60)), scientific=FALSE)` hours to complete.
That, of course, is still quite a lot of time, but not untypical for relatively high-end computational analyses, such as MCMC based Bayesian analyses, that are now quite commonplace in the social and behavioural sciences.

On the other hand, the GPU seems to make a big difference.
The Threadripper without the GPU would still require around `r format(round(mean(gpu_device_cpu_only) * 4 * 409 * 10 / (60 * 60)), scientific=FALSE)` hours, or around `r format(round(mean(gpu_device_cpu_only) * 4 * 409 * 10 / (60 * 60 * 24), 1), scientific=FALSE)` days, to complete.
This is around `r round(mean(gpu_device_cpu_only)/mean(gpu_device), 1)` slower than when the GPU is used and only around `r round(mean(cpu_device)/mean(gpu_device_cpu_only), 1)` faster than the Ryzen 9 despite the fact that the Threadripper has around five times as many CPU cores^[Of course the comparison is not very straightforward because this particular Ryzen 9 model's cores are probably better than those of this particular Threadripper model.
The [cpubenchmark](https://www.cpubenchmark.net/compare/5027vs3837/AMD-Ryzen-9-7900X-vs-AMD-Ryzen-Threadripper-PRO-3995WX) rating puts the Ryzen 9 7900X's single core rating at 4245 compared to 2600 for the Threadripper PRO 3995WX].

Exactly what the best hardware for local LLMs, or rather the best bang-for-the-buck hardware, is not clear from this test.
Would a high-end desktop plus a good GPU be sufficient for efficiently doing a task like the one described in [Part II](part2.html)?
That probably depends largely on what we mean by a "good" GPU?
The RTX A6000 is a relatively expensive GPU.
For example, as of early 2025, retail for around £7500 in the UK.
For that price at that time, you get 18176 CUDA cores and 48GB of ECC VRAM.
Is it necessary to spend this amount or higher?
A proper discussion of that is beyond what we cover here.

# Extending the token memory of Ollama models

One final practical matter is that by default Ollama models (all of them, I believe) have a default token memory of 8192 tokens.
For the task described in [Part II](part2.html) this was sufficient, but when I experimented with similar analyses much larger texts and prompts, specifically texts with over 5000 words and prompts of approximately 1500 this lead to disastrous results.
Only the final parts of the text and some parts (hard to say, but it seemed more like the start only) was being used.
To solve this problem, you need to extend the token memory.
This can not be done on the fly in `ellmer$chat_ollama`.
Instead, you have to make a custom Ollama model.
This is extremely easy to do, however.

You create a text file like the following, which usually named `Modelfile` but it does not have to be:
```bash
FROM llama3.3:latest

PARAMETER num_ctx 65536
```
Then, in the terminal do something like
```bash
ollama create llama3.3x -f Modelfile
```
where the `llama3.3x` is any name of your choice.
This will almost instantaneously create a new LLM, based on Llama3.3, but with a token memory now of 65536 tokens.
Then in R using `ellmer`, you would call this model with `chat_ollama(model = "llama3.3x")`.